Thesis outline

### 2017-06-13 KellyDouglass/Thesis/DirectMethodPhysics.tex
Updated Appendix A

\section{Temperatures from the forbidden emission lines}

@@

Expanding the description in Sec. \ref{sec:O3}, we can define the ratio of the 
emission line strengths for [\ion{O}{3}] $\lambda$4363 and [\ion{O}{3}] 
$\lambda \lambda$4959,5007 as
\begin{equation}
    \frac{j_{4959} + j_{5007}}{j_{4363}} = \frac{\Omega(^3\text{P}, ^1\text{D})}{\Omega(^3\text{P}, ^1\text{S})} \left[ \frac{A_{^1\text{S}, ^1\text{D}} + A_{^1\text{S}, ^3\text{P}}}{A_{^1\text{S}, ^1\text{D}}} \right] \frac{\bar{\nu}(^3\text{P}, ^1\text{D})}{\nu(^1\text{D}, ^1\text{S})} e^{\Delta E/kT}
\end{equation}
where $j_\lambda$ is the emissivity of the emission line, $\Omega (i,j)$ is the 
collision strength between levels $i$ and $j$, $A_{i,j}$ is the radiative 
transition probability between an upper level $i$ and a lower level $j$, $\nu$ 
is the frequency of the transition, $\Delta E$ is the energy difference between 
the $^1$D$_2$ and $^1$S$_0$ levels, and $T$ is the electron temperature 
\cite{Osterbrock89}.  The transition probabilities $A_{i,j}$ do not depend on 
the temperature, but the collision strength $\Omega (i,j)$ is 
temperature-dependent.  We can define
\begin{equation}
    \bar{\nu}(^3\text{P}, ^1\text{D}) = \frac{A_{^1\text{D}_2, ^3\text{P}_2} \nu (5007) + A_{^1\text{D}_2, ^3\text{P}_1} \nu (4959)}{A_{^1\text{D}_2, ^3\text{P}_2} + A_{^1\text{D}_2, ^3\text{P}_1}}
\end{equation}
Inserting numerical values for the collision strengths and transition 
probabilities from \cite{Osterbrock89}, the ratio becomes
\begin{equation}
    \frac{j_{4959} + j_{5007}}{j_{4363}} = \frac{7.73 \exp[(3.29\times 10^4)/T]}{1 + 4.5\times 10^{-4} (N_e/T^{1/2})}
\end{equation}
By measuring the flux of these emission lines, we can then solve for the 
temperature of the gas.

%\section{[\ion{O}{3}]}
%
%There are three significant emission lines for doubly-ionized oxygen.  The 
%relative excitation rates to the $^1$S and $^1$D energy levels depend very 
%strongly on the electron temperature, $T_e$; therefore, the relative strengths 
%of these emitted lines can be used to measure the electron temperature 
%\citep{Osterbrock89}.  In the low-density limit $(n_e < 10^5 \text{ cm}^{-3})$, 
%most excitations to the $^1$D level result in an emission of a photon with a 
%wavelength of either $5007\text{\AA}$ or $4959\text{\AA}$, as shown in Fig. 
%\ref{fig:transitions_P1}.  Most excitations up to $^1$S produce a photon of 
%wavelength $4363\text{\AA}$, followed by a photon of either of the two previous 
%wavelengths (since the electron is now in the $^1$D level).
%
%At higher densities, collisional de-excitation begins to influence these 
%emission rates \citep{Osterbrock89}.  Because the $^1$D level has a longer 
%lifetime than the $^1$S state, it is collisionally de-excited at lower electron 
%densities.  This weakens the $\lambda 4959$ and $\lambda 5007$ emission lines.  
%At the same time, the additional collisional excitations of the $^1$D state 
%permitted by the higher electron densities strengthen the $\lambda 4363$ 
%emission line.
%
%[\ion{O}{3}] $\lambda 4363$ is a temperature-sensitive forbidden transition line 
%of doubly-ionized oxygen that is the preferred line to use when measuring the 
%metallicity of galaxies.  Since the most effective cooling channel in \ion{H}{2} 
%regions is oxygen line emission, lower metallicity regions have higher 
%temperatures \citep{Saintonge07}.  Collisional excitations up to this energy 
%level are more common at higher temperatures, since there are more electrons 
%with the kinetic energy required to excite the O$^{++}$ ion to this energy 
%level.  As a result, the line strength of [\ion{O}{3}] $\lambda 4363$ correlates 
%with the region's temperature and is therefore anticorrelated with the 
%metallicity of the galaxy.

%\section{[\ion{N}{2}]}
%
%The energy-level diagram for the various transitions of [\ion{N}{2}] is very 
%similar to that of [\ion{O}{3}], since they have the same electron ground-state 
%configuration ((1s)$^2$(2s)$^2$(2p)$^2$).  The similarities can be seen in Fig. 
%\ref{fig:transitions_P2}.  Therefore, an estimate of the electron temperature can 
%be made from the [\ion{N}{2}] $\lambda 5755$ emission line.  However, this line 
%is weaker than the [\ion{O}{3}] $\lambda 4363$ auroral line (since there is less 
%N than O in galaxies), so we use the [\ion{O}{3}] auroral line for our 
%temperature estimates, as in Paper I.  After obtaining a temperature and density 
%estimate, we use the [\ion{N}{2}] $\lambda \lambda 6548, 6584$ doublet to 
%estimate the abundance of singly ionized nitrogen in a galaxy.

@@

The relative excitation rates of the two lines shown in Fig. 
\ref{fig:transitions_P1} are proportional to their statistical weights; thus, the 
ratio of the line intensities is a constant in the low-density limit 
\citep{Osterbrock89}.  In the high-density regime, this ratio is best accurately 
described by a Boltzmann population ratio.  There is a critical density for the 
energy levels which describes the turning point between these two extremes.

>>>>

\section{Forbidden transitions}

Similar to a fingerprint, every element has a unique set of wavelengths seen in 
its spectrum that corresponds to the atom's different energy levels.  Since 
electrons within an atom have discrete energy levels, transitions between levels 
result in either the emission (when decreasing in energy) or absorption 
(an increase in energy) of a photon of the same energy as the difference between 
the two levels.  The separation between energy levels is unique for each 
element, so the photon energies either emitted or absorbed by each element are 
unique.

\subsection{Time-dependent perturbation theory}

A time-dependent potential is added to the Schr\"{o}dinger equation in order to 
solve for the forbidden transition energies.  We can assume that the time 
perturbation is a minor component of the potential and can thus treat it as a 
perturbation.  For a simplified two-level system, the time-dependent wave 
function is
\begin{equation}
    \Psi (t) = c_a (t) \psi_a e^{iE_a t/\hbar} + c_b (t) \psi (b) e^{-iE_b t/\hbar}
\end{equation}
To solve for the time-dependent coefficients, we substitute $\Psi (t)$ into the 
time-dependent Schr\"{o}dinter equation
\begin{equation}
    H\Psi = i\hbar \frac{\partial \Psi}{\partial t}, \qquad \text{where } H = H^0 + H'(t)
\end{equation}
We find that
\begin{align}
    \dot{c}_a &= -\frac{i}{\hbar} H'_{ab} e^{-i\omega_0 t} c_b\\
    \dot{c}_b &= -\frac{i}{\hbar} H'_{ba} e^{-i\omega_0 t} c_a
\end{align}
where
\begin{equation}
    \omega_0 \equiv \frac{E_b - E_a}{\hbar}
\end{equation}
and
\begin{equation}
    H_{ij} \equiv \langle \psi_i | H' | \psi_j \rangle
\end{equation}
The probability that transmission has occurred is $P_{b\to a} = |c_a (t)|^2$.  
The transmission rate $R_{a\to b} = dP/dt$.

For an electromagnetic wave, the atom experiences an oscillating electric field.  
For visible light, it is typically safe to assume that the spatial portion of 
the field does not change, since the wavelength is much longer than the size of 
the atom.  However, if we consider a spatial variation of the field, we find 
that the true electric field is
\begin{equation}
    \vec{E}(\vec{r},t) = \vec{E}_0 \cos (\vec{k}\cdot \vec{r} - \omega t)
\end{equation}
Expanding this to first order, we find that
\begin{equation}
    \vec{E}(\vec{r},t) = \vec{E}_0 [\cos (\omega t) + (\vec{k} \cdot \vec{r} \sin (\omega t)]
\end{equation}
The first term in this approximation represents the electric dipole transitions 
(the canonical electron transitions).  The second term leads to forbidden 
transitions (magnetic dipole and electric quadrupole).
%This oscillating field corresponds to our perturbing Hamiltonian:
%\begin{equation}
%    H' = -qE_0 z\cos (\omega t)
%\end{equation}
%where $q$ is the charge on the electron, and the electric field is oscillating 
%in the $z$ direction.  Thus,
%\begin{equation}
%    H'_{ba} = -pE_0 \cos (\omega t), \qquad \text{where } p \equiv q \langle \psi_b |z| \psi_a \rangle
%\end{equation}

%\subsection{Forbidden transitions --- metastable energy levels}

%\subsection{Selection rules}
%
%The transmission rates depend on a matrix element of the form 
%$\langle \psi_b |\vec{r}| \psi_a \rangle$.  Denoting the states by the quantum 
%numbers $n$, $l$, and $m$, the matrix elements can be written as
%\begin{equation}\label{eq:matrix_elements}
%    \langle n'l'm' |\vec{r}| nlm \rangle
%\end{equation}
%By applying the various angular momentum commutators \citep[as shown in][]
%{Griffiths05}, we can develop a set of selection rules for allowed transitions 
%of the electron in an atom.
%
%Inserting the commutation relation $[L_z, z] = 0$ into Eqn. 
%\ref{eq:matrix_elements}, we find that
%\begin{align}
%    0 &= \langle n'l'm'|[L_z, z]|nlm \rangle\\
%    &= \langle n'l'm'|(L_z z - zL_z)|nlm \rangle\\
%    &= \langle n'l'm'|(m'\hbar)z - z(m\hbar)|nlm \rangle\\
%    0 &= (m' - m)\hbar \langle n'l'm'|z|nlm \rangle
%\end{align}
%$\langle n'l'm'|z|nlm \rangle = 0$ is a boring result, so we find that
%\begin{equation}\label{eq:m'=m}
%    m' = m
%\end{equation}
%
%Likewise, by employing the relation $[L_z, x] = i\hbar y$, we find that
%\begin{equation}\label{eq:Lz_x}
%    (m' - m)\hbar \langle n'l'm'|x|nlm \rangle = i\hbar \langle n'l'm'|y|nlm \rangle
%\end{equation}
%With the relation $[L_z, y] = -i\hbar x$, we find that
%\begin{equation}\label{eq:Lz_y}
%    (m' - m)\hbar \langle n'l'm'|y|nlm \rangle = -i\hbar \langle n'l'm'|x|nlm \rangle
%\end{equation}
%Combining Eqns. \ref{eq:Lz_x} and \ref{eq:Lz_y} (and again, ignoring the boring 
%result), we find that
%\begin{equation}\label{eq:m'=+/-m}
%    (m' - m)^2 = 1
%\end{equation}
%Eqns. \ref{eq:m'=m} and \ref{eq:m'=+/-m} show that transitions will only occur 
%if $\Delta m = \pm 1,0$.

\section[Estimating the temperature]{Temperatures from the forbidden emission lines}

@@

Expanding the description in Sec. \ref{sec:O3}, we can define the ratio of the 
emission line strengths for [\ion{O}{3}] $\lambda$4363 and [\ion{O}{3}] 
$\lambda \lambda$4959,5007 as
\begin{equation}
    \frac{j_{4959} + j_{5007}}{j_{4363}} = \frac{\Omega(^3\text{P}, ^1\text{D})}{\Omega(^3\text{P}, ^1\text{S})} \left[ \frac{A_{^1\text{S}, ^1\text{D}} + A_{^1\text{S}, ^3\text{P}}}{A_{^1\text{S}, ^1\text{D}}} \right] \frac{\bar{\nu}(^3\text{P}, ^1\text{D})}{\nu(^1\text{D}, ^1\text{S})} e^{\Delta E/kT}
\end{equation}
where
\begin{equation}
    \bar{\nu}(^3\text{P}, ^1\text{D}) = \frac{A_{^1\text{D}_2, ^3\text{P}_2}\nu (\lambda 5007) + A_{^1\text{D}_2, ^3\text{P}_1}\nu (\lambda 4959)}{A_{^1\text{D}_2, ^3\text{P}_2} + A_{^1\text{D}_2, ^3\text{P}_1}}
\end{equation}
and $j_\lambda$ is the emissivity of the emission line, $\Omega (i,j)$ is the 
collision strength between levels $i$ and $j$, $A_{i,j}$ is the radiative 
transition probability between an upper level $i$ and a lower level $j$, $\nu$ 
is the frequency of the transition, $\Delta E$ is the energy difference between 
the $^1$D$_2$ and $^1$S$_0$ levels, and $T$ is the electron temperature 
\citep{Osterbrock89}.  The transition probabilities $A_{i,j}$ do not depend on 
the temperature, but the collision strength $\Omega (i,j)$ is 
temperature-dependent.  We can define
\begin{equation}
    \bar{\nu}(^3\text{P}, ^1\text{D}) = \frac{A_{^1\text{D}_2, ^3\text{P}_2} \nu (5007) + A_{^1\text{D}_2, ^3\text{P}_1} \nu (4959)}{A_{^1\text{D}_2, ^3\text{P}_2} + A_{^1\text{D}_2, ^3\text{P}_1}}
\end{equation}
Inserting numerical values for the collision strengths and transition 
probabilities from \cite{Osterbrock89}, the ratio becomes
\begin{equation}
    \frac{j_{4959} + j_{5007}}{j_{4363}} = \frac{7.73 \exp[(3.29\times 10^4)/T]}{1 + 4.5\times 10^{-4} (N_e/T^{1/2})}
\end{equation}
Fig. \ref{fig:OIII_temp} shows how this ratio depends on the temperature.  By 
measuring the flux of these emission lines, we can then solve for the 
temperature of the gas.

\begin{figure}
    \includegraphics[width=0.75\textwidth]{Images/Appendix/[OIII]}
    \caption[{[\ion{O}{3}] ratio as a function of temperature}]{The [\ion{O}{3}] 
    $\lambda$4363/($\lambda$4959 + $\lambda$5007) ratio as a function of 
    temperature for various densities \citep{Luridiana15}.  Typical electron 
    densities in \ion{H}{2} regions are $\sim$100 cm$^{-3}$.}
    \label{fig:OIII_temp}
\end{figure}

@@

The relative excitation rates of the two lines shown in Fig. 
\ref{fig:transitions_P1} are proportional to their statistical weights; thus, the 
ratio of the line intensities is a constant in the low-density limit 
\citep{Osterbrock89}.  In the high-density regime, this ratio is best accurately 
described by a Boltzmann population ratio.  There is a critical density for the 
energy levels which describes the turning point between these two extremes.  A 
graphical representation of this ratio as a function of density can be seen in 
Fig. \ref{fig:SII_den}.

\begin{figure}
    \includegraphics[width=0.75\textwidth]{Images/Appendix/[SII]}
    \caption[{[\ion{S}{2}] ratio as a function of density}]{The [\ion{S}{2}] 
    $\lambda$6731/$\lambda$6717 ratio as a function of density for various 
    temperatures \citep{Luridiana15}.  The gas temperatures of \ion{H}{2} 
    regions are typically around 10$^4$ K.}
    \label{fig:SII_den}
\end{figure}
